A closer look at fixed effects regression in structural equation modeling using \texttt{lavaan}

### 2020-07-28 article_format/article_format.tex
Changed the formating, made some minor edits

% Options for packages loaded elsewhere
\PassOptionsToPackage{unicode}{hyperref}
\PassOptionsToPackage{hyphens}{url}
\PassOptionsToPackage{dvipsnames,svgnames*,x11names*}{xcolor}
%
\documentclass[
  12pt,
  a4paper]{article}
\usepackage{lmodern}
\usepackage{amssymb,amsmath}
\usepackage{ifxetex,ifluatex}
\ifnum 0\ifxetex 1\fi\ifluatex 1\fi=0 % if pdftex
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
  \usepackage{textcomp} % provide euro and other symbols
\else % if luatex or xetex
  \usepackage{unicode-math}
  \defaultfontfeatures{Scale=MatchLowercase}
  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
\fi
% Use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\IfFileExists{microtype.sty}{% use microtype if available
  \usepackage[]{microtype}
  \UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\makeatletter
\@ifundefined{KOMAClassName}{% if non-KOMA class
  \IfFileExists{parskip.sty}{%
    \usepackage{parskip}
  }{% else
    \setlength{\parindent}{0pt}
    \setlength{\parskip}{6pt plus 2pt minus 1pt}}
}{% if KOMA class
  \KOMAoptions{parskip=half}}
\makeatother
\usepackage{xcolor}
\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\IfFileExists{bookmark.sty}{\usepackage{bookmark}}{\usepackage{hyperref}}
\hypersetup{
  pdftitle={A closer look at fixed effects regression in structural equation modeling using},
  pdfauthor={Henrik Kenneth Andersen},
  colorlinks=true,
  linkcolor=Maroon,
  filecolor=Maroon,
  citecolor=Blue,
  urlcolor=blue,
  pdfcreator={LaTeX via pandoc}}
\urlstyle{same} % disable monospaced font for URLs
\usepackage[margin=1in]{geometry}
\usepackage{color}
\usepackage{fancyvrb}
\newcommand{\VerbBar}{|}
\newcommand{\VERB}{\Verb[commandchars=\\\{\}]}
\DefineVerbatimEnvironment{Highlighting}{Verbatim}{commandchars=\\\{\}}
% Add ',fontsize=\small' for more characters per line
\usepackage{framed}
\definecolor{shadecolor}{RGB}{248,248,248}
\newenvironment{Shaded}{\begin{snugshade}}{\end{snugshade}}
\newcommand{\AlertTok}[1]{\textcolor[rgb]{0.94,0.16,0.16}{#1}}
\newcommand{\AnnotationTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}
\newcommand{\AttributeTok}[1]{\textcolor[rgb]{0.77,0.63,0.00}{#1}}
\newcommand{\BaseNTok}[1]{\textcolor[rgb]{0.00,0.00,0.81}{#1}}
\newcommand{\BuiltInTok}[1]{#1}
\newcommand{\CharTok}[1]{\textcolor[rgb]{0.31,0.60,0.02}{#1}}
\newcommand{\CommentTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textit{#1}}}
\newcommand{\CommentVarTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}
\newcommand{\ConstantTok}[1]{\textcolor[rgb]{0.00,0.00,0.00}{#1}}
\newcommand{\ControlFlowTok}[1]{\textcolor[rgb]{0.13,0.29,0.53}{\textbf{#1}}}
\newcommand{\DataTypeTok}[1]{\textcolor[rgb]{0.13,0.29,0.53}{#1}}
\newcommand{\DecValTok}[1]{\textcolor[rgb]{0.00,0.00,0.81}{#1}}
\newcommand{\DocumentationTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}
\newcommand{\ErrorTok}[1]{\textcolor[rgb]{0.64,0.00,0.00}{\textbf{#1}}}
\newcommand{\ExtensionTok}[1]{#1}
\newcommand{\FloatTok}[1]{\textcolor[rgb]{0.00,0.00,0.81}{#1}}
\newcommand{\FunctionTok}[1]{\textcolor[rgb]{0.00,0.00,0.00}{#1}}
\newcommand{\ImportTok}[1]{#1}
\newcommand{\InformationTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}
\newcommand{\KeywordTok}[1]{\textcolor[rgb]{0.13,0.29,0.53}{\textbf{#1}}}
\newcommand{\NormalTok}[1]{#1}
\newcommand{\OperatorTok}[1]{\textcolor[rgb]{0.81,0.36,0.00}{\textbf{#1}}}
\newcommand{\OtherTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{#1}}
\newcommand{\PreprocessorTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textit{#1}}}
\newcommand{\RegionMarkerTok}[1]{#1}
\newcommand{\SpecialCharTok}[1]{\textcolor[rgb]{0.00,0.00,0.00}{#1}}
\newcommand{\SpecialStringTok}[1]{\textcolor[rgb]{0.31,0.60,0.02}{#1}}
\newcommand{\StringTok}[1]{\textcolor[rgb]{0.31,0.60,0.02}{#1}}
\newcommand{\VariableTok}[1]{\textcolor[rgb]{0.00,0.00,0.00}{#1}}
\newcommand{\VerbatimStringTok}[1]{\textcolor[rgb]{0.31,0.60,0.02}{#1}}
\newcommand{\WarningTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}
\usepackage{graphicx,grffile}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
% Set default figure placement to htbp
\makeatletter
\def\fps@figure{htbp}
\makeatother
\setlength{\emergencystretch}{3em} % prevent overfull lines
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
\setcounter{secnumdepth}{5}
% ----- Packages

\usepackage{booktabs}
\usepackage{titling}
\usepackage{bm}
\usepackage{mathtools}
\usepackage{amssymb}
\usepackage{amsmath}
\usepackage{subfig}
\usepackage{tikz}
\usetikzlibrary{arrows}
\usepackage[nofiglist]{endfloat}
%\floatsetup[figure]{capposition=top}
%\usepackage{kbordermatrix}
\usepackage{blkarray}
\usepackage{setspace}
\usepackage{etoolbox}
\doublespacing

% ----- Math operators

\DeclareMathOperator{\E}{\mathbb{E}}
\DeclareMathOperator{\Var}{\mathrm{Var}}
\DeclareMathOperator{\Cov}{\mathrm{Cov}}
\DeclareMathOperator*{\argmax}{arg\,max}
\DeclareMathOperator*{\argmin}{arg\,min}

% ----- Graphics

%\graphicspath{{./images/}}

% ----- Equation labels

\mathtoolsset{showonlyrefs}

% ----- Page numbering

\pagenumbering{gobble}

% ----- Spacing

%\renewenvironment{knitrout}{\vspace{10em}}{}

\BeforeBeginEnvironment{equation}{\begin{singlespace}\vspace*{-\baselineskip}}
\AfterEndEnvironment{equation}{\end{singlespace}\noindent\ignorespaces}
\BeforeBeginEnvironment{align}{\begin{singlespace}\vspace*{-\baselineskip}}
\AfterEndEnvironment{align}{\end{singlespace}\noindent\ignorespaces}

\title{\singlespacing A closer look at fixed effects regression in structural
equation modeling using \texttt{lavaan}}
\author{Henrik Kenneth Andersen}
\date{2020-07-27}

\begin{document}
\maketitle
\begin{abstract}
This article provides an in-depth look at fixed effects regression in
the structural equation modeling (SEM) framework, specifically the
application of fixed effects in the \texttt{lavaan} package for
\texttt{R}. It is meant as a applied guide for researchers, covering the
underlying model specification, syntax, and summary output. Online
supplementary materials further discuss various common extentsions to
the basic fixed-effect model, demonstrating how to relax model
assumptions, deal with measurement error in both the dependent and
independent variables, and include time-invariant predictors in a type
of hybrid fixed-/ random effects model. extbf\{Keywords: Fixed effects,
structural equation modeling, lavaan, R, panel analysis\}
\end{abstract}

@@

Several years ago, Curran and Bauer
(\protect\hyperlink{ref-Curran2011}{2011}) reflected positively on the
growing use of panel studies in empirical social research. Some of the
strengths of panel data are well-known, e.g., the ability to establish
temporal precedence, increased statistical power and the reduction of
potential alternative models. However, perhaps the greatest strength of
panel data is that they allow for a more rigorous testing of substantive
theories. Panel data, i.e., repeated measures of the same observed units
(people, schools, firms, countries, etc.), allow researchers to
decompose the error term into a part that stays constant within units
and the part that changes over time. The part that does not change over
time can be seen as the combined effect of all time-invariant influences
(e.g., sex, date of birth, nationality) on the dependent variable. Fixed
effects (FE) regression involves controlling for these time-invariant
influences via a number of various methods. It thus accounts for a
likely and common source of bias.

@@

There are a number of articles describing basic concept of panel model
regression, and FE regression in SEM (e.g., Allison
\protect\hyperlink{ref-Allison2011}{2011}; Bollen and Brand
\protect\hyperlink{ref-Bollen2010}{2010}; Teachman et al.
\protect\hyperlink{ref-Teachman2001}{2001}). This article is intended as
a \textit{practical guide} for researchers looking for in-depth help
with specifying FE models in SEM. It focuses on the \texttt{lavaan}
(Rosseel \protect\hyperlink{ref-R-lavaan}{2012}) package for \texttt{R}
(R Core Team \protect\hyperlink{ref-R-base}{2017}). While \texttt{Mplus}
(Muthén and Muthén, \protect\hyperlink{ref-Mplus}{n.d.}) is arguably the
most robust SEM software currently available (in terms of features like
alignment, latent variable interactions, for example), the
\texttt{lavaan} package has many benefits. First, like \texttt{R} it is
open source and completely free. For researchers dipping their toes into
SEM, there is no financial barrier to try, and no risk if they decide it
is not for them. Second, the implementation of \texttt{lavaan} in the
larger \texttt{R} environment is an enormous advantage. Instead of
poring over reams of plain text, copying out coefficients by hand, every
part of the \texttt{lavaan} output is available as an object. This means
that all aspects of the model, from fit indices, to coefficients and
standard errors, to the model matrices, can be accessed and easily
integrated into tables and plots. Furthermore, \texttt{R} can be used
for a great deal of applications. It can be used to manage and
manipulate as well as simulate data, perform symbolic algebra, run more
traditional analyses (e.g., multiple regression, logistic regression,
principal component analysis), etc. Once one is comfortable using
\texttt{R}, there is no longer any need to switch between different
software for data preparation and analysis.

The following article outlines the basic idea of FE regression, the
particularities of FE in SEM, and shows its implementation in
\texttt{lavaan}. Using
\href{https://github.com/henrik-andersen/FE-SEM/blob/master/simulation-code.R}{simulated
data}, it demonstrates and annotates the code for the most basic FE
model and provides an overview of the summary output. One of the main
strengths of panel SEM compared to the more traditional methods of panel
analysis is its flexibility. Therefore, a number of potential extensions
to the basic model, including relaxing various assumptions, dealing with
measurement error in both the independent and dependent variables, as
well as the inclusion of time-invariant predictors in the form of a
hybrid fixed-/ random effects model, are shown in detail in the form of
\href{https://github.com/henrik-andersen/FE-SEM/blob/master/extensions.pdf}{online
supplementary materials}.

@@

To begin, let us start by reviewing a general panel model (Bollen and
Brand \protect\hyperlink{ref-Bollen2010}{2010}), also referred to as the
`unobserved effects model' (Wooldridge
\protect\hyperlink{ref-Wooldridge2012}{2012}; Croissant and Millo
\protect\hyperlink{ref-R-plm_a}{2008}) (we will return to this model in
the
\href{https://github.com/henrik-andersen/FE-SEM/blob/master/extensions.pdf}{online
supplementary materials} of the article when we discuss loosening
assumptions) \begin{align}
y_{it} & = \bm{x}_{it}\bm{\beta} + \bm{z}_{i}\bm{\gamma} + \alpha_{i} + \varepsilon_{it} \label{eq:gpm}
\end{align} where \(y_{it}\) is the dependent variable for unit
\(i, \ i = 1, ..., N\) at time \(t, \ t = 1, ..., T\), \(\bm{x}_{it}\)
is a \(1 \times K\) vector of time-varying covariates (which could
include a constant) linked to the dependent variable by the
\(K \times 1\) vector of coefficients \(\bm{\beta}\). \(\bm{z}_{i}\) is
a \(1 \times M\) vector of time-invariant covariates linked to the
dependent variable by the \(M \times 1\) vector of coefficients in
\(\bm{\gamma}\), \(\alpha_{i}\) represents the combined effect of all
unobserved time-constant variables affecting the dependent variable and
\(\varepsilon_{it}\) is the idiosyncratic error.

@@

Consistency of the following models requires the assumption of strict
exogeneity, although what constitutes strict exogeneity differs between
the random and fixed effects setups. Each assumption will be discussed
shortly. Apart from that, we typically make the following assumptions
(see, e.g., Wooldridge \protect\hyperlink{ref-Wooldridge2002}{2002};
Schmidheiny \protect\hyperlink{ref-Schmidheiny2019}{2019}) about this
model:

@@

For the random effects (RE) model, we define a composite error term:
\(\nu_{it} = \alpha_{i} + \varepsilon_{it}\) and rewrite the model in
Equation \eqref{eq:gpm} as \begin{align}
y_{it} & = \bm{x}_{it}\bm{\beta} + \bm{z}_{i}\bm{\gamma} + \nu_{it}, \ \text{or} \\
\bm{y}_{i} & = \bm{X}_{i}\bm{\beta} + \bm{Z}_{i}\bm{\gamma} + \bm{\nu}_{i},
\end{align} where
\(\bm{\nu}_{i} = \alpha_{i} \bm{\iota}_{T} + \bm{\varepsilon}_{i}\) and
\(\bm{\iota}_{T}\) is a \(T \times 1\) vector of ones. The strict
exogeneity assumption in the RE model implies \begin{align}
\E[\varepsilon_{it} | \bm{X}_{i}, \bm{z}_{i}, \alpha_{i}] & = 0, \\
\E[\alpha_{i} |\bm{X}_{i}, \bm{z}_{i}] = \E[\alpha_{i}] & = 0,
\end{align} where \(\bm{X}_{i} = \bm{x}_{i1}, ..., \bm{x}_{iT}\). For
both parts, the assumption that the unconditional expectations are 0 is
unproblematic as long as a constant is included in the regression. The
first part says the idiosyncratic errors at each timepoint are assumed
to be independent of the explanatory variables at \textit{all}
timepoints which is stronger than just assuming that they are
\textit{contemporaneously} independent. This implies that they are also
uncorrelated, i.e.,
\(\E[\bm{x}_{is}^{\intercal}\varepsilon_{it}] = \bm{0}\) and
\(\E[\bm{z}_{i}^{\intercal}\varepsilon_{it}] = \bm{0}, \ \forall \ s, t = 1, ..., T\)
(Wooldridge \protect\hyperlink{ref-Wooldridge2002}{2002}; Brüderl and
Ludwig \protect\hyperlink{ref-Bruederl2015}{2015}). We assume the
idiosyncratic errors are further independent of the individual effects,
which implies \(\E[\alpha_{i}\varepsilon_{it}] = 0\).

The second part is the potentially controversial assumption: it states
that the individual effects are uncorrelated with the independent
variables. We can use an example to show why this is often
controversial: If we are interested in the question of whether married
men earn more than unmarried men, then the second part of the strict
exogeneity assumption means that a man's marriage status would have to
be uncorrelated with all the time-invariant characteristics that could
potentially make that man an attractive marriage candidate in the first
place; e.g., looks, personality, family's status, profession, etc.
(Brüderl and Ludwig
\protect\hyperlink{ref-Bruederl2015}{2015}).\footnote{Assuming, for the
  sake of argument, that these characteristics are constant over time.}

From what we have discussed so far, the \(T \times T\) covariance matrix
of the errors
\(\bm{\Omega}_{i} = \E[\bm{\nu}_{i}\bm{\nu}_{i}^{\intercal}]\) can be
constructed. However, the standard random effects model adds the
additional assumptions \begin{alignat}{3}
\E[\varepsilon_{it}^{2} | \bm{X}_{i}, \bm{z}_{i}, \alpha_{i}] & = \E[\varepsilon_{it}^{2}] && = \sigma_{\varepsilon}^{2}, && \ t = 1, ..., T, \\
\E[\varepsilon_{it}\varepsilon_{is}| \bm{X}_{i}, \bm{z}_{i}, \alpha_{i}] & = \E[\varepsilon_{it}\varepsilon_{is}] && = 0, && \ \forall \ t \ne s 
\end{alignat} i.e., the idiosyncratic errors are conditionally
homoscedastic and serially uncorrelated and \begin{alignat}{2}
\E[\alpha_{i}^{2}| \bm{X}_{i},\bm{z}_{i}] & = \E[\alpha_{i}^{2}] && = \sigma^{2}_{\alpha}
\end{alignat} i.e., the individual effects are conditionally
homoscedastic (they are necessarily serially correlated as long as
\(\sigma_{\alpha}^{2} > 0\)) we arrive at the typical random effects
structure of the \(NT \times NT\) matrix \(\bm{\Omega}\):

>>>>

% interactcadsample.tex
% v1.03 - April 2017

\documentclass[]{interact}

\usepackage{epstopdf}% To incorporate .eps illustrations using PDFLaTeX, etc.
\usepackage{subfigure}% Support for small, `sub' figures and tables
%\usepackage[nolists,tablesfirst]{endfloat}% To `separate' figures and tables from text if required

\usepackage{natbib}% Citation support using natbib.sty
\bibpunct[, ]{(}{)}{;}{a}{}{,}% Citation support using natbib.sty
\renewcommand\bibfont{\fontsize{10}{12}\selectfont}% Bibliography support using natbib.sty

\theoremstyle{plain}% Theorem-like structures provided by amsthm.sty
\newtheorem{theorem}{Theorem}[section]
\newtheorem{lemma}[theorem]{Lemma}
\newtheorem{corollary}[theorem]{Corollary}
\newtheorem{proposition}[theorem]{Proposition}

\theoremstyle{definition}
\newtheorem{definition}[theorem]{Definition}
\newtheorem{example}[theorem]{Example}

\theoremstyle{remark}
\newtheorem{remark}{Remark}
\newtheorem{notation}{Notation}

% see https://stackoverflow.com/a/47122900
\usepackage{color}
\usepackage{fancyvrb}
\newcommand{\VerbBar}{|}
\newcommand{\VERB}{\Verb[commandchars=\\\{\}]}
\DefineVerbatimEnvironment{Highlighting}{Verbatim}{commandchars=\\\{\}}
% Add ',fontsize=\small' for more characters per line
\usepackage{framed}
\definecolor{shadecolor}{RGB}{248,248,248}
\newenvironment{Shaded}{\begin{snugshade}}{\end{snugshade}}
\newcommand{\AlertTok}[1]{\textcolor[rgb]{0.94,0.16,0.16}{#1}}
\newcommand{\AnnotationTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}
\newcommand{\AttributeTok}[1]{\textcolor[rgb]{0.77,0.63,0.00}{#1}}
\newcommand{\BaseNTok}[1]{\textcolor[rgb]{0.00,0.00,0.81}{#1}}
\newcommand{\BuiltInTok}[1]{#1}
\newcommand{\CharTok}[1]{\textcolor[rgb]{0.31,0.60,0.02}{#1}}
\newcommand{\CommentTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textit{#1}}}
\newcommand{\CommentVarTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}
\newcommand{\ConstantTok}[1]{\textcolor[rgb]{0.00,0.00,0.00}{#1}}
\newcommand{\ControlFlowTok}[1]{\textcolor[rgb]{0.13,0.29,0.53}{\textbf{#1}}}
\newcommand{\DataTypeTok}[1]{\textcolor[rgb]{0.13,0.29,0.53}{#1}}
\newcommand{\DecValTok}[1]{\textcolor[rgb]{0.00,0.00,0.81}{#1}}
\newcommand{\DocumentationTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}
\newcommand{\ErrorTok}[1]{\textcolor[rgb]{0.64,0.00,0.00}{\textbf{#1}}}
\newcommand{\ExtensionTok}[1]{#1}
\newcommand{\FloatTok}[1]{\textcolor[rgb]{0.00,0.00,0.81}{#1}}
\newcommand{\FunctionTok}[1]{\textcolor[rgb]{0.00,0.00,0.00}{#1}}
\newcommand{\ImportTok}[1]{#1}
\newcommand{\InformationTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}
\newcommand{\KeywordTok}[1]{\textcolor[rgb]{0.13,0.29,0.53}{\textbf{#1}}}
\newcommand{\NormalTok}[1]{#1}
\newcommand{\OperatorTok}[1]{\textcolor[rgb]{0.81,0.36,0.00}{\textbf{#1}}}
\newcommand{\OtherTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{#1}}
\newcommand{\PreprocessorTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textit{#1}}}
\newcommand{\RegionMarkerTok}[1]{#1}
\newcommand{\SpecialCharTok}[1]{\textcolor[rgb]{0.00,0.00,0.00}{#1}}
\newcommand{\SpecialStringTok}[1]{\textcolor[rgb]{0.31,0.60,0.02}{#1}}
\newcommand{\StringTok}[1]{\textcolor[rgb]{0.31,0.60,0.02}{#1}}
\newcommand{\VariableTok}[1]{\textcolor[rgb]{0.00,0.00,0.00}{#1}}
\newcommand{\VerbatimStringTok}[1]{\textcolor[rgb]{0.31,0.60,0.02}{#1}}
\newcommand{\WarningTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}

\usepackage{hyperref}
\usepackage[utf8]{inputenc}
\usepackage{booktabs}
\usepackage{bm}
\usepackage{mathtools}
\usepackage{amssymb}
\usepackage{amsmath}
\usepackage{tikz}
\usetikzlibrary{arrows}
\usepackage[nofiglist]{endfloat}
\usepackage{blkarray}
\usepackage{setspace}
\usepackage{etoolbox}
\DeclareMathOperator{\E}{\mathbb{E}}
\DeclareMathOperator{\Var}{\mathrm{Var}}
\DeclareMathOperator{\Cov}{\mathrm{Cov}}
\DeclareMathOperator*{\argmax}{arg\,max}
\DeclareMathOperator*{\argmin}{arg\,min}
\mathtoolsset{showonlyrefs}
\BeforeBeginEnvironment{equation}{\begin{singlespace}\vspace*{-\baselineskip}}
\AfterEndEnvironment{equation}{\end{singlespace}\noindent\ignorespaces}
\BeforeBeginEnvironment{align}{\begin{singlespace}\vspace*{-\baselineskip}}
\AfterEndEnvironment{align}{\end{singlespace}\noindent\ignorespaces}
\def\tightlist{}
\pagenumbering{gobble}

\begin{document}

\articletype{}

\title{A closer look at fixed effects regression in structural equation
modeling using \texttt{lavaan}}

\author{\name{Henrik Kenneth Andersen$^{a}$}
\affil{$^{a}$Chemnitz University of Technology, Institute of Sociology, Chair for
Empirical Social Research, Thüringer Weg 9, 09126 Chemnitz, Germany}
}

\thanks{CONTACT Henrik Kenneth Andersen. Email: \href{mailto:henrik.andersen@soziologie.tu-chemnitz.de}{\nolinkurl{henrik.andersen@soziologie.tu-chemnitz.de}}}

\maketitle

\begin{abstract}
This article provides an in-depth look at fixed effects regression in
the structural equation modeling (SEM) framework, specifically the
application of fixed effects in the \texttt{lavaan} package for
\texttt{R}. It is meant as a applied guide for researchers, covering the
underlying model specification, syntax, and summary output. Online
supplementary materials further discuss various common extensions to the
basic fixed-effect model, demonstrating how to relax model assumptions,
deal with measurement error in both the dependent and independent
variables, and include time-invariant predictors in a type of hybrid
fixed-/ random effects model.
\end{abstract}

\begin{keywords}
Fixed effects, structural equation modeling, lavaan, R, panel analysis
\end{keywords}

@@

Several years ago, \citet{Curran2011} reflected positively on the
growing use of panel studies in empirical social research. Some of the
strengths of panel data are well-known, e.g., the ability to establish
temporal precedence, increased statistical power and the reduction of
potential alternative models. However, perhaps the greatest strength of
panel data is that they allow for a more rigorous testing of substantive
theories. Panel data, i.e., repeated measures of the same observed units
(people, schools, firms, countries, etc.), allow researchers to
decompose the error term into a part that stays constant within units
and the part that changes over time. The part that does not change over
time can be seen as the combined effect of all time-invariant influences
(e.g., sex, date of birth, nationality) on the dependent variable. Fixed
effects (FE) regression involves controlling for these time-invariant
influences via a number of various methods. It thus accounts for a
likely and common source of bias.

@@

There are a number of articles describing basic concept of panel model
regression, and FE regression in SEM
\citep[e.g.,][]{Allison2011, Bollen2010, Teachman2001}. This article is
intended as a \textit{practical guide} for researchers looking for
in-depth help with specifying FE models in SEM. It focuses on the
\texttt{lavaan} \citep{R-lavaan} package for \texttt{R} \citep{R-base}.
While \texttt{Mplus} \citep{Mplus} is arguably the most robust SEM
software currently available (in terms of features like alignment,
latent variable interactions, for example), the \texttt{lavaan} package
has many benefits. First, like \texttt{R} it is open source and
completely free. For researchers dipping their toes into SEM, there is
no financial barrier to try, and no risk if they decide it is not for
them. Second, the implementation of \texttt{lavaan} in the larger
\texttt{R} environment is an enormous advantage. Instead of poring over
reams of plain text, copying out coefficients by hand, every part of the
\texttt{lavaan} output is available as an object. This means that all
aspects of the model, from fit indices, to coefficients and standard
errors, to the model matrices, can be accessed and easily integrated
into tables and plots. Furthermore, \texttt{R} can be used for a great
deal of applications. It can be used to manage and manipulate as well as
simulate data, perform symbolic algebra, run more traditional analyses
(e.g., multiple regression, logistic regression, principal component
analysis), etc. Once one is comfortable using \texttt{R}, there is no
longer any need to switch between different software for data
preparation and analysis.

The following article outlines the basic idea of panel regression, the
particularities of panel regression SEM, and shows its implementation in
\texttt{lavaan}. The focus on FE panel regression stems from the fact
that the assumptions needed to justify other common models, e.g., random
effects, are often implausible. As we will see, the practical
differences between the two are, in any event, very small. Using
\href{https://github.com/henrik-andersen/FE-SEM/blob/master/simulation-code.R}{simulated
data}, it demonstrates and annotates the code for the most basic FE
model and provides an overview of the summary output. One of the main
strengths of panel SEM compared to the more traditional methods of panel
analysis is its flexibility. Therefore, a number of potential extensions
to the basic model, including relaxing various assumptions, dealing with
measurement error in both the independent and dependent variables, as
well as the inclusion of time-invariant predictors in the form of a
hybrid fixed-/ random effects model, are shown in detail in the form of
\href{https://github.com/henrik-andersen/FE-SEM/blob/master/extensions.pdf}{online
supplementary materials}.

@@

To begin, let us start by reviewing a general panel model
\citep{Bollen2010}, also referred to as the `unobserved effects model'
\citep{Wooldridge2012, R-plm_a} (we will return to this model in the
\href{https://github.com/henrik-andersen/FE-SEM/blob/master/extensions.pdf}{online
supplementary materials} when we discuss loosening assumptions)
\begin{align}
y_{it} & = \bm{x}_{it}\bm{\beta} + \bm{z}_{i}\bm{\gamma} + \alpha_{i} + \varepsilon_{it} \label{eq:gpm}
\end{align} where \(y_{it}\) is the dependent variable for unit
\(i, \ i = 1, ..., N\) at time \(t, \ t = 1, ..., T\), \(\bm{x}_{it}\)
is a \(1 \times K\) vector of time-varying covariates (which could
include a constant) linked to the dependent variable by the
\(K \times 1\) vector of coefficients \(\bm{\beta}\). \(\bm{z}_{i}\) is
a \(1 \times M\) vector of time-invariant covariates linked to the
dependent variable by the \(M \times 1\) vector of coefficients in
\(\bm{\gamma}\), \(\alpha_{i}\) represents the combined effect of all
unobserved time-constant variables affecting the dependent variable and
\(\varepsilon_{it}\) is the idiosyncratic error.

@@

Consistency of the following models requires the assumption of strict
exogeneity, although what constitutes strict exogeneity differs between
the random and fixed effects setups. Each assumption will be discussed
shortly. Apart from that, we typically make the following assumptions
about this model \citep[see, e.g.,][]{Wooldridge2002, Schmidheiny2019}:

@@

Somewhat unintuitively, the modern approach to FE regression treats the
unobserved effect \(\alpha_{i}\) as a \textit{random} variable, rather
than a fixed parameter to be estimated for each of the cross sectional
units \citep{Wooldridge2002}. And in fact, this view carries over to the
way in which FE regression is applied in SEM. So, in order to help
facilitate the discussion on the application of FE-SEM later on, it
makes sense to spend some time discussing the basic idea of random
effects regression (again, as we will see, the practical difference
between the two models is very small).

For the random effects (RE) model, we define a composite error term:
\(\nu_{it} = \alpha_{i} + \varepsilon_{it}\) and rewrite the model in
Equation \eqref{eq:gpm} as \begin{align}
y_{it} & = \bm{x}_{it}\bm{\beta} + \bm{z}_{i}\bm{\gamma} + \nu_{it}, \ \text{or} \\
\bm{y}_{i} & = \bm{X}_{i}\bm{\beta} + \bm{Z}_{i}\bm{\gamma} + \bm{\nu}_{i},
\end{align} where
\(\bm{\nu}_{i} = \alpha_{i} \bm{\iota}_{T} + \bm{\varepsilon}_{i}\) and
\(\bm{\iota}_{T}\) is a \(T \times 1\) vector of ones. The strict
exogeneity assumption in the RE model implies \begin{align}
\mathop{\mathrm{\mathbb{E}}}[\varepsilon_{it} | \bm{X}_{i}, \bm{z}_{i}, \alpha_{i}] & = 0, \\
\mathop{\mathrm{\mathbb{E}}}[\alpha_{i} |\bm{X}_{i}, \bm{z}_{i}] = \mathop{\mathrm{\mathbb{E}}}[\alpha_{i}] & = 0,
\end{align} where \(\bm{X}_{i} = \bm{x}_{i1}, ..., \bm{x}_{iT}\). For
both parts, the assumption that the unconditional expectations are 0 is
unproblematic as long as a constant is included in the regression. The
first part says the idiosyncratic errors at each timepoint are assumed
to be independent of the explanatory variables at \textit{all}
timepoints which is stronger than just assuming that they are
\textit{contemporaneously} independent. This implies that they are also
uncorrelated, i.e.,
\(\mathop{\mathrm{\mathbb{E}}}[\bm{x}_{is}^{\intercal}\varepsilon_{it}] = \bm{0}\)
and
\(\mathop{\mathrm{\mathbb{E}}}[\bm{z}_{i}^{\intercal}\varepsilon_{it}] = \bm{0}, \ \forall \ s, t = 1, ..., T\)
\citep{Wooldridge2002, Bruederl2015}. We assume the idiosyncratic errors
are further independent of the individual effects, which implies
\(\mathop{\mathrm{\mathbb{E}}}[\alpha_{i}\varepsilon_{it}] = 0\).

The second part is the potentially controversial assumption: it states
that the individual effects are uncorrelated with the independent
variables. We can use an intuitive concrete example to show why this is
often controversial: If we are interested in the question of whether
married men earn more than unmarried men, then the second part of the
strict exogeneity assumption means that a man's marriage status would
have to be uncorrelated with all the time-invariant characteristics that
could potentially make that man an attractive marriage candidate in the
first place; e.g., looks, personality, family's status, profession, etc.
\citep{Bruederl2015}.\footnote{Assuming, for the sake of argument, that
  these characteristics are constant over time.}

From what we have discussed so far, the \(T \times T\) covariance matrix
of the errors
\(\bm{\Omega}_{i} = \mathop{\mathrm{\mathbb{E}}}[\bm{\nu}_{i}\bm{\nu}_{i}^{\intercal}]\)
can be constructed. However, the standard random effects model adds the
additional assumptions \begin{alignat}{3}
\mathop{\mathrm{\mathbb{E}}}[\varepsilon_{it}^{2} | \bm{X}_{i}, \bm{z}_{i}, \alpha_{i}] & = \mathop{\mathrm{\mathbb{E}}}[\varepsilon_{it}^{2}] && = \sigma_{\varepsilon}^{2}, && \ t = 1, ..., T, \\
\mathop{\mathrm{\mathbb{E}}}[\varepsilon_{it}\varepsilon_{is}| \bm{X}_{i}, \bm{z}_{i}, \alpha_{i}] & = \mathop{\mathrm{\mathbb{E}}}[\varepsilon_{it}\varepsilon_{is}] && = 0, && \ \forall \ t \ne s 
\end{alignat} i.e., the idiosyncratic errors are conditionally
homoscedastic and serially uncorrelated and \begin{alignat}{2}
\mathop{\mathrm{\mathbb{E}}}[\alpha_{i}^{2}| \bm{X}_{i},\bm{z}_{i}] & = \mathop{\mathrm{\mathbb{E}}}[\alpha_{i}^{2}] && = \sigma^{2}_{\alpha}
\end{alignat} i.e., the individual effects are conditionally
homoscedastic (they are necessarily serially correlated as long as
\(\sigma_{\alpha}^{2} > 0\)). From that, we arrive at the typical random
effects structure of the \(NT \times NT\) matrix \(\bm{\Omega}\):
